Analysis

### 2023-02-20 analysis/analysis.Rmd
edit readme, recreate makefile, simplify analysis scripts

```{r initialize_packages, message=FALSE}
library(morris)
library(broom)
library(knitr)
library(broman)

library(phyloseq)
library(knitr)
library(magrittr)
library(vegan)
library(ggpubr)
library(ggtext)
library(ALDEx2)
library(corncob)
library(ANCOMBC)
library(tidyverse)
library(here)

@@

@@

```{r plot_raw}
ratio <- calc_plot_ratios(fluxes$passage, fluxes$ch4)

ggplot(fluxes, aes(x = passage, y = ch4, color = treat)) +
    stat_smooth(method = 'lm', se = FALSE, formula = 'y ~ x') +
    scale_color_manual(
        name = "Treatment", 
        labels = c('Neutral', 'Positive'), 
        values = c('gray40', 'darkorange2')) +
    geom_jitter(width = 0.1) + 
    labs(x = "Passage Number", 
         y = "Methane Oxidation Rate (-k)") +
    coord_fixed(ratio) +
  scale_x_continuous(labels = 1:5)
```

@@

```{r plot_log10}
ratio_log10 <- calc_plot_ratios(fluxes$passage, fluxes$log_ch4)

(p_response <- ggplot(fluxes, aes(x = passage, y = log_ch4, color = treat)) +
    stat_smooth(method = 'lm', se = FALSE, formula = 'y ~ x') +
    scale_color_manual(
      name = "Treatment", 
      labels = c('Neutral', 'Positive'), 
      values = c('gray40', 'darkorange2')) +
    geom_jitter(#aes(alpha = selected), 
      width = 0.1) + 
        labs(x = "Passage Number", 
                  y = expression(log[10] ~ CH[4] ~ "(-k)")) +
        coord_fixed(ratio_log10) +
  scale_x_continuous(labels = 1:5))

save(p_response, file = here(der_dir, 'response_plot.RData'))

@@

@@

```{r save_model}
save(response_model, compare_response_mods, per_slope, file = here(der_dir, 'response_model.RData'))
```

@@

## Test heritability as mid-offspring on mid-parent

```{r test_treatment}

fit_no_treat <- lm(offspring ~ selected, data = heritability)
kable(tidy(fit_no_treat))
herit_model <- lm(offspring ~ selected * treat, data = heritability)
kable(tidy(herit_model))
kable(tidy(anova(fit_no_treat, herit_model)))
summary(herit_model)

herit_model_n <- 
heritability %>% 
  filter(treat == "n") %>% 
  lm(offspring ~ selected, data = .) %>% 
  tidy()
herit_model_n
herit_model_p <- 
heritability %>% 
  filter(treat == "p") %>% 
  lm(offspring ~ selected, data = .)
summary(herit_model_p)
```

Here, I fit two models. The first is simply offspring flux on selected parent flux, ignoring treatment. There is no correlation `r print_lm(fit_no_treat, 2)`. Next, I added treatment and the interaction between treatment and parental flux. Here, there is a significant correlation between parent and offspring for both the Neutral treatment `r print_lm(herit_model, 2)` and the Positive treatment `r print_lm(herit_model, 4)`. Notably, the sign of the effect is reversed: negative for the Neutral treatment and positive for the Positive treatment.

```{r plot_herit}

n_label <- paste0('Neutral slope = ', myround(tidy(herit_model)[2, 2], 2))
p_label <- paste0('Positive slope = ', myround(tidy(herit_model)[2, 2] + tidy(herit_model)[4, 2], 2))
  ratio <- calc_plot_ratios(heritability$selected, heritability$offspring)
(p_herit <- 
  ggplot(heritability, aes(selected, offspring, color = treat)) +
    geom_point() + 
    # stat_poly_eq() +
    stat_smooth(method = 'lm', se = F) +
    # annotate("text", x = min(heritability$parental), y = max(heritability$offspring), label = p_label, hjust = 0, color = "darkorange2") +
    # annotate("text", x = min(heritability$parental), y = max(heritability$offspring) - 0.04, label = n_label, hjust = 0, color = "gray40") +
    labs(x = expression(Donor ~ log[10] ~ CH[4] ~ "(-k)"), 
         y = expression(atop(Recipient, log[10] ~ CH[4] ~ "(-k)"))) +
    scale_color_manual(name = "Treatment", 
                       labels = c('Neutral', 'Positive'), 
                       values = c('gray40', 'darkorange2')) +
    coord_fixed(1))

save(p_herit, file = here(der_dir, 'herit_plot.RData'))

# ggsave(here(fig_dir, 'fig2.jpg'), p_herit, width = 4, height = 3)
# ggsave(here(fig_dir, 'fig2.svg'), p_herit, width = 4, height = 3)

@@

```{r save_heritability}
save(herit_model, file = here(der_dir, 'herit_model.RData'))
```

```{r}
heritability_each <- read_tsv(here(der_dir, 'heritability_each.tsv'))

fit_no_treat <- lm(offspring ~ selected, data = heritability_each)
kable(tidy(fit_no_treat))
herit_model <- lm(offspring ~ selected * treat, data = heritability_each)
kable(tidy(herit_model))
kable(tidy(anova(fit_no_treat, herit_model)))

herit_model_n_each <- 
heritability_each %>% 
  filter(treat == "n") %>% 
  lm(offspring ~ selected, data = .) %>% 
  tidy()
herit_model_p_each <- 
heritability_each %>% 
  filter(treat == "p") %>% 
  lm(offspring ~ selected, data = .) %>% 
  tidy()

ratio <- calc_plot_ratios(heritability_each$selected, heritability_each$offspring)
(p_herit_each <- 
ggplot(heritability_each, aes(selected, offspring, color = treat)) +
    # theme_classic() +
    # geom_rangeframe(color = 'black') +
    geom_point() + 
    stat_smooth(method = 'lm', se = F) +
    # stat_poly_line(se = T) +
    # stat_poly_eq(aes(label = paste(after_stat(eq.label),
                                 # after_stat(rr.label), sep = "*\", \"*"))) +
    labs(x = expression(atop(Mean ~ Donor, log[10] ~ CH[4] ~ "(-k)")), 
         y = expression(atop(Recipient, log[10] ~ CH[4] ~ "(-k)"))) +
    scale_color_manual(name = "Treatment", 
                       labels = c('Neutral', 'Positive'), 
                       values = c('gray40', 'darkorange2')) +
    coord_fixed(1))

# save(p_herit_each, file = here(der_dir, 'herit_plot.RData'))
# save(herit_model, file = here(der_dir 'herit_model.RData'))
# 
# ggsave(here(fig_dir, 'fig2_each.jpg'), p_herit_each, width = 4, height = 3) 
# ggsave(here(fig_dir, 'fig2_each.svg'), p_herit_each, width = 4, height = 3) 
```

@@

min_seqs <- paste(substr(as.character(min_seqs), 1, 3), substr(as.character(min_seqs), 4, 6), sep = ",")
save(kt, wt, rare_median, s5, min_seqs, file = here(der_dir, "richness_models.RData"))

@@

centroids <-
  nmds_data %>% 
  group_by(passage, treat) %>% 
  summarize(NMDS1 = mean(NMDS1),
            NMDS2 = mean(NMDS2),
            .groups = "drop")
p_beta <-
  nmds_data %>%   
  ggplot(aes(x = NMDS1, y = NMDS2, shape = passage, color = treat)) +
  geom_point() +
  stat_ellipse(show.legend = FALSE) +
  coord_fixed() +
  scale_shape_discrete(name = "Passage", labels = c(2, 5)) +
  scale_color_manual(name = "Treatment",
                       labels = c('Neutral', 'Positive'),
                       values = c('gray40', 'darkorange2')) #+
  # theme_bw() +
  # theme(
  #   plot.background = element_blank(),
  #   panel.grid.major = element_blank(),
  #   panel.grid.minor = element_blank())

save(p_beta, file = here(der_dir, "beta_fig.RData"))

ggsave(here(fig_dir, 'fig3.png'), p_beta, height = 2.5, width = 4)

ggsave(here(fig_dir, 'fig3.svg'), p_beta, height = 2.5, width = 4)

@@

save(tidy_permanova, file = here(der_dir, "beta_model.RData"))

@@

save(dbrda_aov, dbrda_percent_explained, file = here(der_dir, "dbrda.RData"))

@@

save(p_log10, da_fams, file = here(der_dir, "da_fig.RData"))

ggsave(here(fig_dir, 'fig4.jpg'), p_log10, width = 5, height = 5) 
ggsave(here(fig_dir, 'fig4.svg'), p_log10, width = 5, height = 5)

@@

>>>>

```{r initialize_packages, message=FALSE}
library(morris)
library(broom)
library(knitr)
library(broman)
library(phyloseq)
library(knitr)
library(magrittr)
library(vegan)
library(ggpubr)
library(ggtext)
library(ALDEx2)
library(corncob)
library(ANCOMBC)
library(tidyverse)
library(here)
library(patchwork)

@@

```{r}
aes_data %>% 
    select(sample, count, Phylum, Family) %>% 
    group_by(sample) %>% 
    mutate(sum_count = sum(count)) %>% 
    mutate(ra = count / sum_count) %>% 
    filter(is.na(Family)) %>%
    # filter(Family == "") %>%
    summarize(sum = sum(ra))

@@

```{r plot_raw}

ggplot(fluxes, aes(x = passage, y = ch4, color = treat)) +
    stat_smooth(method = 'lm', se = FALSE, formula = 'y ~ x') +
    scale_color_manual(
        name = "Treatment", 
        labels = c('Neutral', 'Positive'), 
        values = c('gray40', 'darkorange2')) +
    geom_jitter(width = 0.1) + 
    labs(x = "Passage Number", 
         y = "Methane Oxidation Rate (-k)") +
  scale_x_continuous(labels = 1:5)
```

@@

```{r plot_log10}

ratio = calc_plot_ratios(fluxes$passage, fluxes$log_ch4)

(p_response <- ggplot(fluxes, aes(x = passage, y = log_ch4, color = treat)) +
    stat_smooth(method = 'lm', se = FALSE, formula = 'y ~ x') +
    scale_color_manual(
      name = "Treatment", 
      labels = c('Neutral', 'Positive'), 
      values = c('gray40', 'darkorange2')) +
    geom_jitter(#aes(alpha = selected), 
      width = 0.1) + 
        labs(x = "Passage Number", 
                  y = expression(log[10] ~ CH[4] ~ "(-k)")) +
  scale_x_continuous(labels = 1:5) +
    coord_fixed(ratio))

# save(p_response, file = here(der_dir, 'response_plot.RData'))

@@

Since the residuals are not normally distributed, we will use the log-transformed data.

@@

```{r save_model}
# save(response_model, compare_response_mods, per_slope, file = here(der_dir, 'response_model.RData'))
```

@@

## Test heritability as divergence on cumulative selection differential

```{r heritability}

divergence <-
    fluxes %>% 
    select(passage, treat, log_ch4) %>% 
    group_by(passage, treat) %>% 
    summarize(mean = mean(log_ch4), .groups = "drop") %>% 
    pivot_wider(names_from = "treat", values_from = "mean") %>% 
    mutate(divergence = p - n, .keep = 'unused')

# Selection Differential
sd <- 
    heritability %>% 
    mutate(sd = selected - parental) %>% 
    select(passage, treat, sd) %>% 
    pivot_wider(names_from = 'treat', values_from = 'sd') %>% 
    mutate(sd_sum = p + n, .keep = 'unused')

sd %>% 
    arrange(passage) %>% 
    mutate(sd_csum = cumsum(sd_sum)) ->
    sd_csum

# Realized Heritability
rh <-
    divergence %>% 
    left_join(sd_csum, by = "passage") %>% 
    filter(passage != 0 )
```

ratio_div <- calc_plot_ratios(divergence$passage, divergence$divergence)
(
p_div <-
ggplot(divergence, 
       aes(x = passage, 
           y = divergence)) +
    stat_smooth(method = 'lm', 
                se = FALSE, 
                formula = 'y ~ x',
                color = "black",
                linewidth = 0.5) +
    geom_point() + 
    labs(x = expression(atop("Passage Number", " ")), 
         y = expression(atop("Divergence", log[10] ~ CH[4] ~ "(-k)"))) +
    coord_fixed(ratio_div) +
    scale_x_continuous(labels = 1:5) + 
    scale_y_continuous(n.breaks = 4) +
    theme(plot.margin = margin(0, 0.25, 0, 0.5, "cm"))
)
ratio_rh <- calc_plot_ratios(rh$sd_csum, rh$divergence)

(
p_rh <-
rh %>% 
ggplot(aes(x = sd_csum, 
           y = divergence)) +
    stat_smooth(method = 'lm',
                se = F,
                formula = 'y ~ x',
                color = "black",
                linewidth = 0.5) +
    geom_point() + 
    labs(x = expression(atop("Cumulative Selection Differential", log[10] ~ CH[4] ~ "(-k)")), 
         y = expression(atop("Divergence", log[10] ~ CH[4] ~ "(-k)"))) +
    scale_y_continuous(n.breaks = 4) +
    # theme(plot.margin = margin(0, 0.5, 0, 0.25, "cm")) +
    coord_fixed()
)

remove_y <- theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(),
    axis.title.y = element_blank()
)
p <- list(p_div, p_rh)
herit_plot <- wrap_plots(p, nrow = 1) + plot_layout(guides = "collect")

rh_model<-lm(divergence~sd_csum,data = rh)
summary(rh_model)

# save(herit_plot, file = here(der_dir, "herit_plot.RData"))

# ggsave(here(fig_dir, "fig2.png"), width = 8*.9, height = 4*.9)
# ggsave(here(fig_dir, "fig2.svg"), width = 8*.9, height = 4*.9)

# save(rh_model, file = here(der_dir, 'herit_model.RData'))

@@

@@

min_seqs <- paste(substr(as.character(min_seqs), 1, 3), substr(as.character(min_seqs), 4, 6), sep = ",")
# save(kt, wt, rare_median, s5, min_seqs, file = here(der_dir, "richness_models.RData"))

@@

centroids <-
  nmds_data %>% 
  group_by(passage, treat) %>% 
  summarize(NMDS1 = mean(NMDS1),
            NMDS2 = mean(NMDS2),
            .groups = "drop")
p_beta <-
  nmds_data %>%   
  ggplot(aes(x = NMDS1, y = NMDS2, shape = passage, color = treat)) +
  geom_point() +
  stat_ellipse(show.legend = FALSE) +
  coord_fixed() +
  scale_shape_discrete(name = "Passage", labels = c(2, 5)) +
  scale_color_manual(name = "Treatment",
                       labels = c('Neutral', 'Positive'),
                       values = c('gray40', 'darkorange2')) #+
  # theme_bw() +
  # theme(
  #   plot.background = element_blank(),
  #   panel.grid.major = element_blank(),
  #   panel.grid.minor = element_blank())
  #   
pcoa <- cmdscale(rare_dist_bray, eig = TRUE, add = TRUE)
round(pcoa$eig*100/sum(pcoa$eig),1)
pcoa_per_exp <- 100 * eigenvals(pcoa) / sum(eigenvals(pcoa))

ratio = pcoa_per_exp[2]/pcoa_per_exp[1]

pcoa_scores <-
  scores(pcoa, display = "sites") %>% 
  as_tibble(rownames = "sample")

pcoa_data <-
  aes_data %>% 
  select(sample, passage, treat) %>% 
  distinct %>% 
  inner_join(pcoa_scores, by = "sample")
# p_beta <-
  pcoa_data %>%   
  ggplot(aes(x = Dim1, y = Dim2, shape = passage, color = treat)) +
  geom_point() +
  stat_ellipse(show.legend = FALSE) +
  coord_fixed(ratio) +
  scale_shape_discrete(name = "Passage", labels = c(2, 5)) +
  scale_color_manual(name = "Treatment",
                       labels = c('Neutral', 'Positive'),
                       values = c('gray40', 'darkorange2')) #+
  # theme_bw() +
  # theme(
  #   plot.background = element_blank(),
  #   panel.grid.major = element_blank(),
  #   panel.grid.minor = element_blank())

# save(p_beta, file = here(der_dir, "beta_fig.RData"))

# ggsave(here(fig_dir, 'fig3.png'), p_beta, height = 2.5, width = 4) 
# ggsave(here(fig_dir, 'fig3.svg'), p_beta, height = 2.5, width = 4)

@@

# save(tidy_permanova, file = here(der_dir, "beta_model.RData"))

@@

# save(dbrda_aov, dbrda_percent_explained, file = here(der_dir, "dbrda.RData"))

@@

# save(p_log10, da_fams, file = here(der_dir, "da_fig.RData"))

# ggsave(here(fig_dir, 'fig4.jpg'), p_log10, width = 5, height = 5) 
# ggsave(here(fig_dir, 'fig4.svg'), p_log10, width = 5, height = 5)

@@

```{r save_outputs}

save(p_response, herit_plot, p_beta, p_log10, file = here(der_dir, 'plots.RData'))
save(response_model, compare_response_mods, per_slope, rh_model, kt, wt, rare_median, s5, min_seqs, tidy_permanova, dbrda_aov, dbrda_percent_explained, da_fams, file = here(der_dir, 'models.RData'))

# save(p_response, file = here(der_dir, 'response_plot.RData'))
# save(response_model, compare_response_mods, per_slope, file = here(der_dir, 'response_model.RData'))
# 
# save(herit_plot, file = here(der_dir, "herit_plot.RData"))

# save(rh_model, file = here(der_dir, 'herit_model.RData'))
# save(kt, wt, rare_median, s5, min_seqs, file = here(der_dir, "richness_models.RData"))
# save(p_beta, file = here(der_dir, "beta_fig.RData"))
# save(tidy_permanova, file = here(der_dir, "beta_model.RData"))
# save(dbrda_aov, dbrda_percent_explained, file = here(der_dir, "dbrda.RData"))
# save(p_log10, da_fams, file = here(der_dir, "da_fig.RData"))
